---
title: "p8105_hw2_nck2132"
author: "Nina Knitowski"
date: "2026-10-07"
output: github_document
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Libraries
```{r Data SetUp}
library(tidyverse)

NYC_Transit_Data= read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
NYC_Transit_Data = janitor::clean_names(NYC_Transit_Data)
names(NYC_Transit_Data)
```

#Question 1

Select Only Necessary Columns
```{r Necessary Columns}
NYC_Transit_Data <- NYC_Transit_Data |>
select( 
line, station_name, station_latitude,station_longitude, starts_with("route"),
       vending, entry,entrance_type, ada_compliance=ada)
```
Convert Entry 
```{r Convert Entry}
NYC_Transit_Data = 
 NYC_Transit_Data |>
  mutate("entry" = case_match(
      entry,
      "YES" ~ TRUE,
      "NO"  ~ FALSE
    ))
```
Describe Data 
```{r View}
head(NYC_Transit_Data)
```
The dataset contains variables on NYC MTA station details and their compliance.

For data cleaning, the dataset now contains only the columns `r names(NYC_Transit_Data)`. We would still want to combine all the routes into one column "routes served" to fully tidy the data.`line`, `station_name`, `entrance_type`, and `vending` are character variables. `Entry` and `ada_compliance` are logical variables. `station_lattitude` and `station_longitude` are numerical variables. `route1`, `route2`, `route3`, `route4`, `route4`, `route5`, `route6`, `route7` are character variables and `route8`,`route9`, `route10`, `route11` are numerical variables. This data is not fully tidy yet especially the routes 1-9 variables these should be combined and be the same variable type. The data has `r nrow(NYC_Transit_Data)` rows, `r ncol(NYC_Transit_Data)` columns.

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.
```{r}
n_stations <- NYC_Transit_Data |> 
  distinct(station_name, line) |> 
  nrow()
```

Based on the output, there are `r n_stations` distinct stations in New York City MTA.

How many stations are ADA compliant?
```{r ADA_Complaint}
n_stations_ada <- NYC_Transit_Data |> 
  distinct(line, station_name, .keep_all = TRUE) |> 
  filter(ada_compliance == TRUE) |> 
  nrow()
```
There are `r n_stations_ada` stations that are ada_compliant in NYC MTA system. 

What proportion of station entrances / exits without vending allow entrance?
```{r Prop}
prop_vending <- NYC_Transit_Data |> 
   filter(vending == "NO") |> 
  summarize(proportion = mean(entry == TRUE)) |> 
  pull(proportion)
``` 

`r round(prop_vending * 100, 1)`% of station entrances/exits without vending allow entrance.

Tidyy Data for Route Name vs. Number
```{r Convert}
NYC_Transit_Data_tidy = NYC_Transit_Data |> 
  mutate(across(starts_with("route"), as.character)) |>
  pivot_longer( 
    route1:route11,
    names_to = "route_number", 
    values_to = "route_name")
values_drop_na = TRUE
```

Check it worked
```{r}
head(NYC_Transit_Data_tidy)
```

```{r Distinct Stations A Train}
distinctA_stations <- NYC_Transit_Data_tidy |>
  filter(route_name == 'A') |> 
   distinct(line, station_name) |> 
  nrow()
```
There are `r distinctA_stations` distinct stations that service the A train.

```{r Prop A Train ADA Compliant}
propA_compliant <-NYC_Transit_Data_tidy |>
  filter(route_name == 'A', ada_compliance =="TRUE") |> 
   distinct(line, station_name) |> 
  nrow()
```
Of those `r propA_compliant` distinct stations that service the A train, 17 are ADA compliant.

#Question 2 

Load Mr. Trash Wheel 
```{r Data Setup}
library(readxl)
Mr.Trash_df = read_excel(
  path = "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel",
  range = "A2:N742",
  skip = 1,
  na = c(".", "NA", ""),
)
Mr.Trash_df = janitor::clean_names(Mr.Trash_df)
```

Cleaning Mr. Trash Wheel
```{r Sports_Ball Convert}
Mr.Trash_df = Mr.Trash_df |>
  filter(!is.na(dumpster)) |>
  mutate(sports_balls = as.integer(round(sports_balls, digits = 0)), 
         year = as.numeric(year))
  mutate(Mr.Trash_df, trash_wheel_type = "Mr. Trash")
```

Load Professor Trash Wheel Data & Clean
```{r Professor Trash Wheel Data}
Professor_Trash_Wheel_df = read_excel(
    path = "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
range = "A2:M139",
col_types = c(
    "numeric", "text", "numeric", "date", "numeric", "numeric",
    "numeric", "numeric", "numeric", "numeric", "numeric", "numeric",
    "numeric"
 ),
skip = 1,
na = c(".", "NA", ""),
)
Professor_Trash_Wheel_df = janitor::clean_names(Professor_Trash_Wheel_df)

Professor_Trash_Wheel_df = Professor_Trash_Wheel_df |>
mutate(trash_wheel_type = "Professor", 
       year = as.numeric(year))
```

```{r Gwynns Falls Data}
Gwynns_Falls_Trash_Wheel_df = read_excel(
    path = "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = 4,
    range = "A2:K394",
    skip = 1,
    na = c(".", "NA", "")
)
Gwynns_Falls_Trash_Wheel_df = janitor::clean_names(Gwynns_Falls_Trash_Wheel_df)

Gwynns_Falls_Trash_Wheel_df = mutate(Gwynns_Falls_Trash_Wheel_df,trash_wheel_type = "Gwynns Falls",
       year = as.numeric(year)
       )
```

```{r Combine Tidy Datasets}
combined_trash_wheels = bind_rows(Mr.Trash_df,Gwynns_Falls_Trash_Wheel_df, Professor_Trash_Wheel_df)
nrow(combined_trash_wheels)
```
```{r Total Weight}
prof_totalweight=
  combined_trash_wheels |>
  filter(trash_wheel_type == "Professor")|>
  pull(weight_tons) |>
  sum(na.rm=TRUE)
prof_totalweight
```


```{r Total Cigarette Butts }
gwynnda_2022cig= 
  combined_trash_wheels|>
  filter(trash_wheel_type == "Gwynns Falls", month=="June", year== "2022") |>
  pull(cigarette_butts) |>
  sum(na.rm=TRUE)
gwynnda_2022cig
```

The combined dataset called "combined_trash_wheels" and the dataset contains data from "Mr. Trash Wheel", "Professor Trash Wheel", and "Gwynns Falls Trash Wheel." The combined dataset contains `r nrow(combined_trash_wheels)` observations across `r ncol(combined_trash_wheels)` variables. In "Professor Trash Wheel" dataset there were `r prof_totalweight` tons of trash and in "Gwynna Falls Trash Wheel" there were `r gwynnda_2022cig` of total cigarette butts collect. 

#Question 3
Import MCI_baseline & Tidy Data
```{r Import, Tidy}
MCI_Df=read_csv(
  "./Data/data_mci/MCI_baseline.csv",
    skip=1,
  na = "."
) |> 
  janitor::clean_names() |>
 mutate(
    sex   = case_match(sex, 0 ~ "female", 1 ~ "male") |> factor(),
    apoe4 = case_match(apoe4, 0 ~ "non-carrier", 1 ~ "carrier") |> factor()
  )
```

Cleaning Data P2, Descriptive Data
```{r Eligible Description}
mci_eligible = MCI_Df |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

n_eligible = nrow(mci_eligible)

n_develop_mci = mci_eligible |>
  filter(!is.na(age_at_onset)) |>
  nrow()

mean_baseline_age = mci_eligible |>
  summarize(mean_age = mean(current_age)) |>
  pull(mean_age)

prop_women_apoe4 = mci_eligible |>
  filter(sex == "female") |>
  summarize(prop = mean(apoe4 == "carrier")) |>
  pull(prop)
```

I imported the baseline data using `skip = 1` to drop the description row and `na = "."` to code missing values. I then recoded sex and APOE4 carrier status so the values are labeled ("female"/"male" and "carrier"/"non-carrier") instead of 0 and 1. Participants who already had MCI at baseline were removed, leaving `r n_eligible` eligible participants. Of these, `r n_develop_mci` developed MCI during follow-up. The average baseline age was `r round(mean_baseline_age, 2)` years, and `r round(100 * prop_women_apoe4, 1)`% of women in the study were APOE4 carriers.

Data Import MCI_Amyloid and Clean
```{r} 
mci_amyloid_df =
  read_csv(
    "./Data/data_mci/mci_amyloid.csv",
    skip = 1,
    na = c("NA", "Na")
  ) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |>
  mutate(
    time = replace(time, time == "baseline", "0"),
    time = as.numeric(time)
  )

print(mci_amyloid_df, n = 12)
```

```{r Amyloid Description}
n_amyloid_obs = nrow(mci_amyloid_df)

n_amyloid_participants = mci_amyloid_df |>
  distinct(id) |>
  nrow()

n_amyloid_missing = mci_amyloid_df |>
  filter(is.na(amyloid_ratio)) |>
  nrow()

amyloid_range = mci_amyloid_df |>
  summarize(min = min(amyloid_ratio, na.rm = TRUE),
            max = max(amyloid_ratio, na.rm = TRUE))
```

The amyloid file has a description row above the real column names, so I imported it with `skip = 1` and coded both "NA" and "Na" as missing, then cleaned the names and renamed the identifier to `id`. I pivoted the data from wide to long so each row is one participant at one visit, with `time` as numeric years since baseline (0, 2, 4, 6, 8). The tidy dataset has `r n_amyloid_obs` observations from `r n_amyloid_participants` participants, with amyloid ratios ranging from about `r round(pull(amyloid_range, min), 3)` to `r round(pull(amyloid_range, max), 3)`. `r n_amyloid_missing` observations are missing, mostly at follow-up visits, so the data are unbalanced, and the file has 487 participants versus 483 in the baseline file. This means there is a slight mismatch in the datasets

Check if Overlap/Unbalanced
```{r check_overlap}
baseline_only = anti_join(MCI_Df, mci_amyloid_df, by = "id") |>
  distinct(id)

amyloid_only = anti_join(mci_amyloid_df, MCI_Df, by = "id") |>
  distinct(id)

baseline_only
amyloid_only

n_baseline_only = nrow(baseline_only)
n_amyloid_only  = nrow(amyloid_only)
```

```{r combine}
mci_combined_df =
  inner_join(MCI_Df, mci_amyloid_df, by = "id")

n_combined = mci_combined_df |>
  distinct(id) |>
  nrow()

n_combined_missing = mci_combined_df |>
  filter(is.na(amyloid_ratio)) |>
  nrow()

write_csv(mci_combined_df, "./Data/data_mci/mci_combined.csv")
```

`r n_baseline_only` participants appear only in the baseline dataset and `r n_amyloid_only` appear only in the amyloid dataset. The files do not say why, but participants missing from the amyloid data may have been enrolled without a biomarker measurement or withdrawn early, while those missing from the baseline data may lack intake records or come from a different recruitment source, so they have no demographic or APOE4 information. I used an inner join on `id` to keep only the `r n_combined` participants present in both files, which gives one dataset linking baseline characteristics (age, sex, education, APOE4 status, age at MCI onset) to amyloid ratios measured at 0, 2, 4, 6 and 8 years after baseline. The combined dataset has `r nrow(mci_combined_df)` rows, one per participant per visit, and was exported as `mci_combined.csv`. It is not balanced: `r n_combined_missing` amyloid values are still missing because some participants skipped follow-up visits, so this will need to be accounted for in later longitudinal analysis.
  

