---
title: "p8105_hw2_nck2132"
author: "Nina Knitowski"
date: "2026-10-07"
output: github_document
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Data SetUp

```{r Data SetUp}
library(tidyverse)
NYC_Transit_Data= read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
NYC_Transit_Data = janitor::clean_names(NYC_Transit_Data)
names(NYC_Transit_Data)
```

#Question 1

Select Only Necessary Columns
```{r Necessary Columns}
NYC_Transit_Data <- NYC_Transit_Data |>
select( 
line, station_name, station_latitude,station_longitude, starts_with("route"),
       vending, entry,entrance_type, ada_compliance=ada)
```
Convert Entry 
```{r Convert Entry}
NYC_Transit_Data = 
 NYC_Transit_Data |>
  mutate("entry" = case_match(
      entry,
      "YES" ~ TRUE,
      "NO"  ~ FALSE
    ))
```
Describe Data 
```{r View}
head(NYC_Transit_Data)
```
The dataset now contains only the columns line, station, name, station latitude / longitude, routes 1-9, entry, vending, entrance type, and ADA compliance. We would still want to combine all the routes into one column "routes served." Line, Station Name, Entrance Type, and Vending are all character variables. Entry and Ada compliance are logical variables. Station Lat/Long is a numerical variable. Routes 1-11 are a mix of number and character variables. This data is not fully tidy yet especially the routes 1-9 variables these should be combined and be character or numerical variables. The dimensions are 1868 X 19 with 19 columns and 1868 obs.

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.
```{r}
NYC_Transit_Data |> 
  distinct(station_name, line) |> 
  nrow()
```

Based on the output, there are 465 distinct stations in New York City MTA.

How many stations are ADA compliant?
```{r ADA_Complaint}
NYC_Transit_Data |> 
  distinct(line, station_name, .keep_all = TRUE) |> 
  filter(ada_compliance == TRUE) |> 
  nrow()
```
There are 84 stations that are ada_compliant in NYC MTA system. 

What proportion of station entrances / exits without vending allow entrance?
```{r Prop}
NYC_Transit_Data |> 
  filter(vending == "NO") |> 
  count(entry) |> 
  mutate(proportion = n / sum(n))
``` 
37.77% of station entrances/exits without vending allow entrance

Tiday Data for Route Name vs. Number
```{r Convert}
NYC_Transit_Data_tidy = NYC_Transit_Data |> 
  mutate(across(starts_with("route"), as.character)) |>
  pivot_longer( 
    route1:route11,
    names_to = "route_number", 
    values_to = "route_name")
values_drop_na = TRUE
```

Check it worked
```{r}
head(NYC_Transit_Data_tidy)
```

```{r Distinct Stations A Train}
NYC_Transit_Data_tidy |>
  filter(route_name == 'A') |> 
   distinct(line, station_name) |> 
  nrow()
```
There are 60 distinct stations that service the A train.

```{r Prop A Train ADA Compliant}
NYC_Transit_Data_tidy |>
  filter(route_name == 'A', ada_compliance =="TRUE") |> 
   distinct(line, station_name) |> 
  nrow()
```
Of those 60 distinct stations that service the A train, 17 are ADA compliant.

#Question 2 

```{r Data Set Up}
library(readxl)
Mr.Trash_df = read_excel(
  path = "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel",
  range = "A2:N741",
  skip = 1,
  na = c(".", "NA", ""),
)
Mr.Trash_df = janitor::clean_names(Mr.Trash_df)
```
```{r Sports_Ball Convert}
Mr.Trash_df = Mr.Trash_df |>
  mutate(sports_balls = as.integer(round(sports_balls, digits = 0)), 
         year = as.numeric(year))
 
mutate(Mr.Trash_df, trash_wheel_type = "Mr. Trash")
```

```{r Professor Trash Wheel Data}
Professor_Trash_Wheel_df = read_excel(
    path = "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel",
range = "A2:L134",
skip = 1,
na = c(".", "NA", ""),
)
Professor_Trash_Wheel_df = janitor::clean_names(Professor_Trash_Wheel_df)

Professor_Trash_Wheel_df = Professor_Trash_Wheel_df |>
mutate(trash_wheel_type = "Professor", 
       year = as.numeric(year))
```

```{r Gwynns Falls Data}
Gwynns_Falls_Trash_Wheel_df = read_excel(
    path = "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = 4,
    range = "A2:K351",
    skip = 1,
    na = c(".", "NA", "")
)
Gwynns_Falls_Trash_Wheel_df = janitor::clean_names(Gwynns_Falls_Trash_Wheel_df)

Gwynns_Falls_Trash_Wheel_df = mutate(Gwynns_Falls_Trash_Wheel_df,trash_wheel_type = "Gwynns Falls",
       year = as.numeric(year)
       )
```

```{r Combine Tidy Datasets}
combined_trash_wheels = bind_rows(Mr.Trash_df,Gwynns_Falls_Trash_Wheel_df, Professor_Trash_Wheel_df)
nrow(combined_trash_wheels)
```
```{r Total Weight}
prof_totalweight=
  combined_trash_wheels |>
  filter(trash_wheel_type == "Professor")|>
  pull(weight_tons) |>
  sum(na.rm=TRUE)
prof_totalweight
```


```{r Total Cigarette Butts }
gwynnda_2022cig= 
  combined_trash_wheels|>
  filter(trash_wheel_type == "Gwynns Falls", month=="June", year== "2022") |>
  pull(cigarette_butts) |>
  sum(na.rm=TRUE)
gwynnda_2022cig
```


The combined dataset called "combined_trash_wheels" and the dataset contains data from "Mr. Trash Wheel", "Professor Trash Wheel", and "Gwynns Falls Trash Wheel." The combined dataset contains `r nrow(combined_trash_wheels)` observations across `r ncol(combined_trash_wheels)` variables. In "Professor Trash Wheel" dataset there were `r prof_totalweight` tons of trash and in "Gwynna Falls Trash Wheel" there were `r gwynnda_2022cig` of total cigarette butts collect. 

